Supplementary Information

### 2020-08-16 doc/manuscript/MEE_manuscript.Rmd
worked on draft

**Traditional methods characterize only linear climate responses, potentially missing important nonlinearities that have been widely observed at other spatio-temporal scales.**
Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2014], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

@@

Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree. We  also ran models with a spline of DBH (3 knots).
Across all possible models, variable AICc weights were summed to determine which drivers are most important. We considered the model containing those variables as the top model.

@@

Temperature responses tended to be negative at lower latitudes, positive at higher.

**Non-linear responses to climate variables were common (Fig. 3).**
- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**based on early results-- need to re-evaluate**)

>>>>

**Traditional methods characterize only linear climate responses, potentially missing important nonlinearities that have been widely observed at other spatio-temporal scales.**
Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2004], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

@@

Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. 1). 
*(DBH not included in models focused on climate sensitivity.)*
We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree. 
Wealso ran models with a spline of DBH (3 knots).
Across all possible models, variable AICc weights were summed to determine which drivers are most important. 
We considered the model containing those variables as the top model.

@@

Precipitation responses were observed at all sites and *for all but # species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
Responses were most commonly positive and were most pronounced at the driest site (Little Tesque). 
...
Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.

Temperature responses were observed at all sites *(but BCI? [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig. 3).
Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this? [ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
